---
title: "Municipal Governance and Voter Turnout: Analyzing the Impacts of City Managers"
author: Noah Ochital^[American University]
format: 
  pdf:
    number-sections: false
    toc: false
    geometry: 
      - top=3cm
editor: source
documentclass: article
date: "`r Sys.Date()`"
abstract: "Municipal elections have long been studied, regarding the large, documented disparity between national and local electoral turnout. These studies have attempted to describe how turnout is correlated with various demographic, institutional, and time-related factors, to identify strategies for increasing local turnout. This paper narrows the focus to one of these predictors: whether a city has a full-time manager. Prior studies have been limited by a lack of data, however, the recently published municipal elections database from Benedictis-Kessner and colleagues enables this analysis to occur. Combining this data, with U.S. Census estimates, and originally collected city manager data, I analyzed a random sample of 396 elections using several regression models. I improved the models’ estimates further by pre-processing the data with a non-parametric matching algorithm to reduce confounding, which refined the sample to 194 elections. I find that despite significant differences in turnout between cities with and without managers, this difference diminishes after controlling the other covariates and becomes statistically insignificant after the matching process."
execute:
  echo: false
  warning: false
  cache: TRUE
  message: false
fig-cap-location: top
tbl-cap-location: top
fig-dpi: 300
fig-pos: "H"
bibliography: ref_govt_310.bib
---

\footnotesize Key Words: Voter Turnout, City Managers, Municipal Elections, Regression Analysis, Propensity Score Matching, Political Participation. \normalsize

\newpage

# Municipal Turnout in Contemporary Politics

In the United States, municipal elections have consistently been shown to have far lower levels of turnout compared to federal elections [@morlan_municipal_1984; @bullock_iii_turnout_1990; @alford_voting_1968]. This divergence is measured as far back as the 1950s, where Bullock (1990) finds in a data set of elections from 1956-1979 , that the average turnout rate in federal elections is 59.1% compared to 31% for municipal elections. This trend is particularly confusing given municipal governments' vast power over their residents’ daily lives, overseeing essential utility, safety, and sanitary services. Additionally, municipalities have control over zoning, building permits, and other regulations which can have numerous and far more tangible impacts on communities than federal or state regulations. Understanding these vast powers, it might make sense to see municipal elections garnering higher turnout rates than federal elections, as potential changes to local governance could affect access to these vital services, but this is not the case. 

Low municipal turnout is not just confusing from a logical perspective, but also has enormousness ramifications for the legitimacy of our political system. The U.S. is built on a system of representative democracy and political participation, but when participation is extremely low, most especially at the most responsive level of government, it indicates that voters do not see any value coming from that system. This is especially dangerous because it allows for officials to be elected with far fewer votes than would should be needed to win a specific office, giving them power which could be abused. Moreover, without enough voters casting ballots it is increasingly difficult for a government to claim the proper mandate to govern. Ideally, even if federal turnout is declining [@bullock_iii_turnout_1990], local turnout should still be at a similar level. However, the deliberate choice by some voters to vote in federal elections but not in local elections [@hajnal_municipal_2003;@caren_big_2007;@morlan_municipal_1984] reflect that there are fundamental deficiencies in the structure or administration of local governments that discourage voters from taking the effort to vote. These are issues that need to be rectified for local governments to regain their legitimacy and be able to accurately represent their constituents. 

Researchers have proposed various factors to explain low municipal turnout typically split into categories involving the timing of elections, overall demographics, and institutional aspects such as government type and electoral systems. Specific factors include but are not limited to, the relative size and demographic components of municipalities, the power of government officials, and the timing of local races compared to national or state races [@hajnal_municipal_2003; @caren_big_2007; @karnig_decline_1983; @kelleher_political_2004; @wood_voter_2002]. This paper narrows the focus to just one factor: the presence of a city manager. City managers are unelected positions with wide-reaching powers and responsibilities as both the chief executive and chief administrator of the city. Although an individual city’s charter sets the powers of the city manager and can vary from city to city, they typically include discretion over the budgets, policies, and administration. Conversely, in cities without managers, these responsibilities must be handled by elected officials, giving them more policy-setting power. City managers can be implemented to any form of municipal governance structure, meaning that cities with a mayor-city council, department commissioner, or just city council structure can all have managers. What distinguishes these managed cities from their unmanaged counterparts is the relative share of power held by the elected officials. 
This shift of power away from elected officials, towards appointed ones, in managed cities, reduces the direct influence voters can have on policies, providing a plausible explanation for decreased levels of turnout [@alford_voting_1968; @hajnal_municipal_2003]. If people think that their votes will not have a sizable impact on their lives or resolve any issues they have with municipal governance because the major decision maker is not elected, then they simply won't vote. Therefore, if this relationship between city manager's and turnout is true, then the removal of the manager form of government would be a simple and effective policy to increase turnout and thereby increasing the legitimacy of local governmental institutions. Many other factors that influence turnout, are more up to free choice, chance, or are difficult to affect with policy, meaning that a critical examination of city manager's provides an important avenue for real and relatively quick change.

The goal of this paper is to estimate the impact that city managers have on municipal turnout, while also updating existing literature with more recent analysis. Although many studies already document clear negative associations between turnout and city managers, these studies either are outdated, or have been limited by lack of data [@hajnal_municipal_2003; @caren_big_2007; @wood_voter_2002]. I plan to update this literature, using a recently published municipal elections database [@benedictis-kessner_american_2023], allowing for a broader sample and more robust analysis. I conducted my analysis by creating several regression models to capture the relationship between city managers and turnout while accounting for covariates, and a matching procedure [@ho_matchit_2024; see @ho_matching_2007] to account for these covariates even further. I hypothesize city managers will negatively impact turnout as previous research describes. However, after performing my analysis, I find that the relationship is not statistically significant.

# Examining the Historical Relationship Between City Managers and Turnout


City Managers were introduced to municipalities in the early 20th century as a central part of governmental reforms, intended to break up powerful partisan, political machines that had controlled many large cities for decades [@bridges_textbook_1997; @caren_big_2007]. Political machines maintained power through corruption and clientelism, giving favors such as monetary assistance or jobs to their voters. Introducing city managers to handle these responsibilities instead of elected officials could potentially cut off the stream of favors, giving fewer incentives for voters to turn out. Another dimension of these reforms was also racial in nature. Many machines had gotten their power from large ethnic minority groups within these cities, so dismantling the machines would return power to white middle-class voters instead [@bridges_textbook_1997].

Early research into the topic has shown that these reforms were remarkably effective at achieving their goals. Several studies examining the decline in municipal turnout document that cities with managers exhibit significantly lower turnout than cities without managers [@alford_voting_1968; @karnig_decline_1983]. Alford and Lee in their 1968 paper (1968), used data from the municipal elections in 1961 and 1962 in cities with populations of 25,000 and concluded that city managers had strong negative correlations with voter turnout, even after accounting for socioeconomic factors such as education and foreign-born percentage. Additionally, Karnig and Walter (1983), found similar results using data from 1935, 1962, and 1975 while also showing that switching away from city managers was positively correlated with voter turnout. This study is particularly important because it’s able to draw data from 1935, which is the oldest of any study, and the closest there is to the adoption of these styles of government.

Bridges (1997) by contrast uses a case study approach examining several cities that were not reformed such as New York and Chicago to what are deemed big-reform cities in the Southwest such as Albuquerque and Phoenix. She shows in her analysis that reformed cities had far lower turnout rates than the non-reformed cities, concluding that in cities such as New York where reform coalitions (and the white middle class) had less power, the level of barriers they were able to erect was far smaller in magnitude than big reform cities [@bridges_textbook_1997]. Although the sample size was small, and the statistical framework weak, the study nonetheless exhibits a strong negative correlation and provides important explanations behind the behavior of the reforms.

Recent research continues to corroborate these claims while using more advanced statistical techniques. In several papers, researchers Caren (2007), Hajnal and Lewis (2003), and Wood (2002) use more recent samples of elections. Wood uses a sample of 57 cities, and elections from 1993-2000, attempting to showcase the relationships between municipal governments on a scale of administrative to political. He finds a clear negative relationship between levels of administration and turnout on his scale, such that more administrative cities with less elected positions have lower turnout [@wood_voter_2002]. Hajnal and Lewis (2003), use a moderate-sized sample of cities from only California, in an advanced ordinary least squares multiple regression that includes many specific institutional characteristics such as mayoral term lengths, veto power, budget authority, and term limits, along with including demographic factors, and electoral concurrency to examine the impact of city managers. Caren (2007) on the other hand, adopted an approach that sampled 38 historically large cities and used a GLS random effects model with similar covariates to Hajnal and Lewis such as having a manager, demographic, and population characteristics, and concurrency, but was able to include non-partisan elections as well. As far as the measurements were concerned all three studies generated statistically significant results, with Wood [@wood_voter_2002] measuring a coefficient of -3%, Hajnal and Lewis [@hajnal_municipal_2003] measuring about -8%, and Caren [@caren_big_2007] measuring -7.5% in their respective models.

The body of research on this topic is very small compared to other areas of political science research, and there were not any articles that concluded the opposite of the studies above. Although all these articles clearly show that city manager's associated with decreases in turnout, the body of information is not large enough to estimate a proper magnitude for the relationship or enough to conclude outright about its direction. Additionally, each of the more recent studies suffer from significant issues that impact the interpretation more generally. Caren’s (2007) study uses a small sample of cities with historically large populations potentially missing how this relationship could change between tiny and large cities. Hajnal and Lewis’ (2003) study only cities from California, very few of which do not have managers due to municipal reforms. This makes the relationship they show hard to believe even given statistical significance, because the coefficient is being determined by small number of cities relative to the sample size. Wood’s (2002) study also showcases a small sample size but is made worse by the five different types of governments used in the study, meaning the number of cities in each group is even smaller and harder to make inferences from. In this study, I attempt to rectify these issues by using a nationwide sample of cities that are diverse in institutional and demographic characteristics, while also contributing to the body as a whole with my findings. 

```{r}
library(tidyverse)
library(tidymodels)
library(stargazer)
library(MatchIt)
library(cobalt)
library(gridExtra)
library(grid)
library(maps)
```

```{r}
#| output: false
cities <- read_csv("../data/cities_constituency_data.csv")
elections <- readRDS("../data/ledb_candidatelevel.rds")
set.seed(435464564)
sample <- sample_n(cities, 150, replace=FALSE)
cities <- cities %>% 
  mutate(sampled = if_else(geo_name %in% sample$geo_name, 1, 0))
elections <- elections %>%
  mutate(sampled = if_else(fips %in% sample$fips , 1 , 0))

sample <- read_csv("../data/sample with manager.csv")
sample <- sample %>%
  mutate(across(c(council.manager, percent_women, percent_white,
                  percent_black, percent_hispanic),as.numeric))

elections$fips <- as.numeric(elections$fips)
sampled_elections <- left_join(sample, elections, by = "fips") %>%
  rename(city = geo_name.x) %>%
  select(fips, state, city, pres_pctD_08, pres_pctD_16, pres_pctD_20, council.manager, state_abb, office_consolidated, district, contest, votes, year, month, incumbent, vote_share) %>%
  filter(year > 2009)
nrow(sampled_elections)

sampled_contests <- sampled_elections %>%
  group_by(contest) %>%
  reframe(
    total.votes = sum(votes),
    year = year,
    city = city,
    office = office_consolidated,
    state = state,
    pres_pctD_08 = pres_pctD_08,
    pres_pctD_16 = pres_pctD_16,
    pres_pctD_20 = pres_pctD_20,
    council.manager = council.manager,
    month = month,
    incumbent = if_else(all(is.na(incumbent)), NA_real_, max(incumbent, na.rm = TRUE)),
    margin = {
      sorted_votes <- sort(votes, decreasing = TRUE)
      if(length(sorted_votes) < 2) {
        100
      }
      else {
      100*(sorted_votes[1] - sorted_votes[2])/(sorted_votes[1]+sorted_votes[2])
      }
    }
  ) %>% 
  filter(!duplicated(contest)) %>%
  filter(grepl("at-large", contest) | grepl("mayor", contest))

pop.2010 <- read_csv("../data/2010.csv")
pop.2011 <-read_csv("../data/2011.csv")
pop.2012 <-read_csv("../data/2012.csv")
pop.2013 <-read_csv("../data/2013.csv")
pop.2014 <-read_csv("../data/2014.csv")
pop.2015 <-read_csv("../data/2015.csv")
pop.2016 <-read_csv("../data/2016.csv")
pop.2017 <-read_csv("../data/2017.csv")
pop.2018 <-read_csv("../data/2018.csv")
pop.2019 <-read_csv("../data/2019.csv")
pop.2020 <-read_csv("../data/2020.csv")
pop.2021 <-read_csv("../data/2021.csv")

p <- function(x) {
  x %>% 
    rename(pop.estimate = DP05_0018E, 
           percent_men = DP05_0002PE,
           percent_women = DP05_0003PE,
           percent_white = DP05_0059PE,
           percent_black = DP05_0060PE,
           percent_hispanic = DP05_0066PE,
           percent_asian_american = DP05_0062PE) %>% 
    mutate(across(c(pop.estimate, percent_men, percent_women, percent_white, 
                    percent_black, percent_hispanic, percent_asian_american),
                  as.numeric)) %>%
    separate(NAME, into = c("city", "state"), sep = ",") %>%
    select(city, state, pop.estimate, percent_men, percent_women, percent_white, 
           percent_black, percent_hispanic, percent_asian_american) %>%
    drop_na()
}

s <- function(x) {
  x %>% 
    rename(pop.estimate = DP05_0021E, 
           percent_men = DP05_0002PE,
           percent_women = DP05_0003PE,
           percent_white = DP05_0064PE,
           percent_black = DP05_0065PE,
           percent_hispanic = DP05_0071PE,
           percent_asian_american = DP05_0067PE) %>% 
    mutate(across(c(pop.estimate, percent_men, percent_women, percent_white, 
                    percent_black, percent_hispanic, percent_asian_american), 
                  as.numeric)) %>%
    separate(NAME, into = c("city", "state"), sep = ",") %>%
    select(city, state, pop.estimate, percent_men, percent_women, percent_white, 
           percent_black, percent_hispanic, percent_asian_american) %>%
    drop_na()
}

t <- function(x, y) {
  x$city <- tolower(x$city)
  y$city <- tolower(y$city)
  y$state <- trimws(y$state)
  left_join(x,y, by= c("city", "state"))
}

poplist <- list(pop.2010, pop.2011, pop.2012, pop.2013, pop.2014, pop.2015,
                pop.2016, pop.2017, pop.2018, pop.2019, pop.2020, pop.2021)


sampled_contests_list <- vector("list", length = 12)

contest_by_year <- split(sampled_contests, sampled_contests$year)


for(i in 1:12) {
  if(i <= 7) {
    poplist[[i]] <- p(poplist[[i]]) 
  } else {
    poplist[[i]] <- s(poplist[[i]])  
  }
  sampled_contests_list[[i]] <- t(contest_by_year[[i]], poplist[[i]])
}

sampled_contests <- bind_rows(sampled_contests_list)

final_contests <- sampled_contests %>%
  mutate(turnout = (total.votes/pop.estimate)*100) %>%
  filter(turnout <= 100) %>% drop_na()

final_contests <- final_contests %>%
  mutate(
    log.pop = log(pop.estimate),
    concurrent = factor(
      case_when(
        year %% 4 == 0 & month == 11 ~ "president",
        year %% 4 == 2 & month == 11 ~ "midterm",
        year %% 4 == 3 & month == 11 & state %in% c("Louisiana") ~ "midterm",
        year %% 4 ==1 & month == 11 & state %in% c("New Jersey", "Virginia") 
        ~ "midterm",
        TRUE ~ "none"), levels = c("none", "midterm", "president")),
    gubernatorial = if_else(
      year %% 4 == 0 & month == 11 & state %in% c("Missouri", "Indiana",
                                                  "North Carolina") |
        year %% 4 == 3 & month == 11 & state %in% c("Louisiana") |
        year %% 4 == 1 & month == 11 & state %in% c("New Jersey", "Virginia") |
        year %% 4 == 2 & month == 11 & state %in% c(
          "Georgia", "Maine", "Michigan", "Oregon", 
          "Arizona", "Rhode Island", "Texas", 
          "Arkansas", "California", "Connecticut", 
          "Florida", "Illinois", "Iowa", "Kansas", 
          "Massachusetts", 
          "North Carolina", "South Carolina", "Ohio", "Pennsylvania", 
          "Colorado", "Minnesota", "Tennessee"),1,0),
    concurrent_pres = if_else(concurrent=="president", 1, 0),
    concurrent_mid = if_else(concurrent=="midterm",1,0),
    office = factor(office, levels = c("City Council", "Mayor")),
    mayor_elec = if_else(office=="Mayor",1,0),
    meanD = mean(c(pres_pctD_08,pres_pctD_16,pres_pctD_20)))

addition <- final_contests %>% group_by(city) %>% summarize(
  has_council = if_else(any(office=="City Council"),1,0),
  has_mayor = if_else(any(office=="Mayor"),1,0),
  same_months = if_else(all(month[office=="City Council"] %in% 
                              month[office=="Mayor"]),1,0 ),
  .groups = "drop") %>%
  filter(has_mayor & has_council) %>% select(same_months, city)

has_mayor <- final_contests %>% group_by(city) %>% summarize(
  has_mayor = if_else(any(office=="Mayor"),1,0), .groups = "drop")

final_contests <- left_join(final_contests, addition, by = "city")

final_contests <- left_join(final_contests, has_mayor, by = "city")

rm(pop.2010,pop.2011,pop.2012, pop.2013,pop.2014,pop.2015,pop.2016,
   pop.2017,pop.2018,pop.2019,pop.2020,pop.2021,poplist,sampled_contests_list,
   contest_by_year)
gc()
```

```{r}
#| output: false

formula <- council.manager ~ concurrent+ has_mayor+ office+ log.pop+ incumbent + 
  margin + percent_women+ percent_black+percent_hispanic+ percent_asian_american
matching_methods <- c("nearest", "optimal", "full", "quick", "genetic")
distance_methods <- c("glm", "gam", "gbm", "lasso", "ridge", "elasticnet",
              "rpart", "randomforest", "nnet", "cbps", 
              "bart", "mahalanobis", "scaled_euclidean", "robust_mahalanobis")
balances <- list()
matching <- function(method, formula, data, cutpoints = NULL, grouping = NULL, 
                     subclass = NULL, distance, distance.options = NULL) {
  match <- matchit(formula, method = method, data = data, cutpoints = cutpoints, 
                   distance = distance, distance.options = distance.options) 
  bal <- bal.tab(match, stats = c("mean.diffs", "variance.ratios"), 
                 binary = "std", thresholds = c(m=0.1, v=2), un = TRUE)
  return(bal$Balance)
}


set.seed(125)
nnet <- list(size = 2)
for(methods in matching_methods) {
  for(distance in distance_methods) { 
    if(distance == "nnet") {
      set.seed(125)
      balances[[paste(methods, distance, sep = "_")]] <- matching(methods, 
    formula, final_contests, distance = distance, distance.options = nnet)
    }
    else {
    set.seed(125)
    balances[[paste(methods, distance, sep = "_")]]<- matching(methods, 
    formula, final_contests, distance = distance)
    }
  }
}
bal_result <- list()
for(i in seq_along(balances)) {
  m <- sum(if_else(balances[[i]]["M.Threshold"]=="Balanced, <0.1",1 , 0), 
           na.rm=TRUE)
  v <- sum(if_else(balances[[i]]["V.Threshold"]=="Balanced, <2", 1 , 0), 
           na.rm=TRUE)
  bal_result[[i]] <-tibble(method = paste(names(balances[i])), 
                    bal_mean = m,
                    bal_var = v)

}
bal_result <- bind_rows(bal_result)
my_pick <- bal_result %>% filter(bal_mean==max(bal_mean)) %>%
  filter(bal_var == max(bal_var)) %>% print()

balances[["optimal_glm"]]
balances[["optimal_gam"]]
tie_breaker <- tibble(
  model = c("optimal_glm", "optimal_gam"),
  un_balanced_mean_diff = 
    c(mean(c(abs(balances[["optimal_glm"]][7,4])-0.1, 
           abs(balances[["optimal_glm"]][11,4])-0.1)),
       mean(c(abs(balances[["optimal_gam"]][7,4])-0.1,
              abs(balances[["optimal_gam"]][11,4])-0.1))))
```

```{r}
set.seed(125)
optimal_glm <- matchit(formula, method = "optimal", data = final_contests, distance = "glm")
matched_data <- match.data(optimal_glm)

```

# Data and Methods

## Data Description

This paper uses a new municipal elections database compiled by Benedictis-Kessner and colleagues in 2023 (2023) containing local election data from 1989-2021. I focus on the subset of the data covering mayoral and town council elections, as these were the only municipal government positions in the dataset. I drew a simple random sample without replacement of `r nrow(sample)` cities representing `r round(mean(cities$sampled)*100, 2)`% of the dataset. I chose this sample size to minimize my original data collection, given the time constraints, while having a large enough sample of elections. QQ-plots in @fig-qq, show the similarity in distributions of key continuous variables between the sample and larger dataset, indicating my sample is representative of the rest of the data.

```{r}
test.vars <- c("population_2020", "mass_ideology_2020","pres_pctD_20",
                    "pres_pctD_16", "pres_pctD_08",
                    "percent_women", "percent_white", "percent_black", "percent_hispanic", 
                    "percent_asian_american") 
qq_labels <- c("Population (2020)", "Ideology Measure (2020)", 
               "Dem. Vote 2020 (%)", 
               "Dem. Vote 2016  (%)",
               "Dem. Vote 2008  (%)",
               "Women (%)", "White (%)", "Black (%)", "Hispanic (%)", 
               "Asian (%)")
quantiles <- list()
qqplots <- list()

for (i in seq_along(test.vars)) {
  sample_q <- quantile(sample[[test.vars[i]]], probs = seq(0.01, 1, 0.01), 
                       na.rm = TRUE)
  pop_q <- quantile(cities[[test.vars[i]]], probs = seq(0.01, 1 , 0.01), 
                    na.rm = TRUE)
  quantiles[[i]] <- tibble(sample_q, pop_q, variable = qq_labels[i])
}

quantiles <- bind_rows(quantiles)
```

```{r}
#| fig.cap: "QQ Plots for Sampled vs. Population Cities"
#| fig.width: 9
#| fig.height: 6
#| label: fig-qq
ggplot(quantiles, aes(sample_q, pop_q)) +
  geom_point()+
  geom_abline(slope = 1, intercept = 0, linetype = 2)+
  labs(x = "Sample Quantiles", y = "Population Quantiles")+
  facet_wrap(~variable, scale = "free", ncol =3)+
  theme_minimal()+
  theme(strip.text = element_text(size = 12),
        axis.text= element_text(size=10),
        axis.title = element_text(size=15))
  

```

I then determined whether each city had managers or not, from their websites. Cities with managers were coded as $1$, without managers coded as $0$, and unincorporated cities as NA. I combined the sample with U.S. Census Data [@u.s.censusbureau; @u.s.censusbureaua], to calculate the turnout rate (using voting age population), and racial demographics. Only the Census data from 2010 and onward was available for download, so I cut off the sample there. Using the sampled cities, I pulled the corresponding data from the elections database, and transformed it to appropriately change the unit of observation to each election, instead of each candidate. After the transformation, `r nrow(sampled_contests) - nrow(final_contests)` elections contained turnout rates over 100%, and were removed. The whole process resulted in `r nrow(final_contests)` elections from `r sum(!is.na(unique(final_contests$city)))` different municipalities. The final list of cities and a map of their locations are in the appendix at @tbl-city and @fig-map respectively.

Summary statistics for the final sample are above in @tbl-sum, all continuous variables except for population are measured in percentages from 1-100, while population is measured on the log scale as $ln(\text{population})$, to account for its large range. All other variables are binary dummy variables, where the mean represents the proportion of elections with that characteristic. Key facts are that the average turnout rate was `r round(mean(final_contests$turnout),0)`%, with a standard deviation of `r round(sd(final_contests$turnout),0)`%, while `r round(mean(final_contests$council.manager),0)`% of elections were in cities with managers indicating a large variance in turnout with a skew in the sample towards cities with managers. It is important to note that this class imbalance may hinder that analysis and all results should be taken carefully.

```{r}
#| output: 'asis'
#| tbl-cap: "Summary Statistics of Selected Covariates in Unmatched Sample"
#| label: tbl-sum
final_contests %>% select(turnout, council.manager, incumbent,margin, 
                          percent_women, percent_black , percent_hispanic , 
                          percent_asian_american, log.pop, concurrent_pres,
                          concurrent_mid,has_mayor,mayor_elec) %>%
  rename("Turnout(%)" = turnout,
         "Has a Manager" = council.manager,
         "Has a Mayor" = has_mayor, 
         "Mayoral Election" = mayor_elec, 
         Incumbency = incumbent,
         "Population(ln)" = log.pop,
         "Margin of Victory(%)" = margin,
         "Women(%)" = percent_women,
         "Black(%)" = percent_black,
         "Asian(%)" = percent_asian_american,
         "Hispanic(%)" = percent_hispanic,
         "Concurrent Presidential" = concurrent_pres,
         "Concurrent Midterm" =  concurrent_mid) %>%
  as.data.frame() %>%
  stargazer(summary = TRUE, header = FALSE, digits = 2,
          single.row = TRUE, column.sep.width = "0pt", font.size = "footnotesize",
          float = FALSE,  omit.summary.stat = "N")
```

## Model Building

The key covariates in my model have been drawn from what other researchers have used in prior research [@hajnal_municipal_2003; @caren_big_2007], however, I was limited by what I could calculate from the Census estimates and electoral data. I was unable to gather data on non-partisan elections nor go into depth on mayoral power, but overall I constructed an expansive dataset with enough predictors to be produce a useful model. The covariates are categorized into three types: institutional (government structure and electoral competitiveness), demographic (population and racial percentages), and electoral timing (concurrency with other elections). Including these additional factors in the model allows for the differences in each city to be accounted for when estimating the impact of a city manager on the turnout rate, creating a better estimate that holds equal other predictors. If these were not included the model would simply attribute the whole difference in turnout rate between the cities with and without managers to be the impact of the manager, which is unrealistic because the city managers are not randomized among municipalities. This is especially important because these factors can be both related to the turnout rate and whether or not a city has a manager.

The main outcome or dependent variable is voter turnout (percentage, 1-100), and the primary predictor is whether the city has a manager (1 = yes, 0 = no). Other factors include margin of victory (percentage, 1-100; uncontested election is a margin of 100), incumbency (1 = incumbent present, 0 = no incumbent), the contested office as a factor with 2 levels (city council, and mayoral), population size (natural log of Census estimate), demographic estimates (sex and ethnicity percentages, 1-100), and concurrency as a factor with 3 levels (not concurrent, midterm concurrency, and presidential concurrency) that tracks concurrency with federal elections. Elections were determined to be concurrent if they occurred within the same year and month as a federal election. Presidential concurrency means the municipal elections took place concurrently with the presidential election. Midterm means the election was held at the same time as the Congressional midterm elections, while all others were labeled not concurrent. Gubernatorial elections are not accounted because most take place during the mid-term cycle, however, gubernatorial elections in New Jersey, Louisiana, and Virginia occur in years without midterm or presidential elections and were thus coded as mid-term, to align them with other states.

Several variables that were present in the dataset were excluded from my estimation of the model, because ordinary least squares(OLS) regression models require that variables cannot be perfectly collinear. The percentage of white residents was strongly correlated with black residents (`r round(cor(final_contests$percent_white, final_contests$percent_black), 2)`), and the percentage of males which was perfectly correlated with women (`r round(cor(final_contests$percent_men, final_contests$percent_women),2)`). For example, when the percent of men increases, the percent of women decreases by the same amount, so computing a coefficient for both is impossible. Therefore for each of these cases, one variable was removed (White(%) and Male(%)), to allow for the estimation of the model. The information contained by these variables is not lost, but instead held by the other demographic variables used in the model. On the other hand, moderately correlated variables such as mayoral election type and mayoral government (`r round(cor(final_contests$mayor_elec, final_contests$has_mayor),2)`) were kept, because they reflect key information about the city and election type necessary for the model. Correlation Matrices for each group are available in the appendix as @tbl-icor, @tbl-dcor, and @tbl-tcor, respectively.

I determined the polynomial degree for each continuous covariate using scatter plots to compare a linear approximation with LOESS estimation provided by `geom_smooth()` in `ggplot2`. As shown in @fig-lm below, although the relationships are not linear, the linear approximation is good at capturing the general trend and intercepts, so have been chosen to reduce the risk of over-fitting the data.

```{r}
#| fig.cap: "Model Building: Comparing Linear Terms to Loess Regression"
#| fig.height: 5
#| fig.width: 8
#| label: fig-lm
final_covs <- final_contests[, c("log.pop", 
                       "margin", "percent_women","percent_black", 
                       "percent_hispanic", "percent_asian_american")]
final_cov_lab <- c("Population (ln)", "Margin of Victory", "Women(%)", "Black (%)",
                   "Hispanic(%)", "Asian(%)")
coef_plots <- list()
for (i in seq_along(final_covs)) {
 coef_plots[[i]]<- ggplot(final_contests, 
                          aes_string(names(final_covs[i]), "turnout"))+
    geom_point()+
   theme_minimal()+ 
   labs(x = paste(final_cov_lab[i]),  y = "Turnout(%)")+
    geom_smooth(method = "lm", se = FALSE)+
   geom_smooth(se = FALSE)
}
grid.arrange(grobs = coef_plots, ncol = 3)
```

The final linear model I estimate is below using turnout as the dependent variable, and the additional measured covariates as independent variables: \begin{multline*}
\text{Turnout}_i = \beta_0 + \beta_1 (\text{Manager}_i) + \beta_2 (\text{Concurrency}_i) +
\beta_3 (\text{Mayoral Government}_i) +\\ \beta_4 (\text{Office of Election}_i) + \beta_5 (\text{Incumbent Present}_i) + 
\beta_6 (\text{Margin of Victory}_i) +\\ \beta_7 (\text{Percent Women}_i) + \beta_8 (\text{Percent Black}_i) +
\beta_9 (\text{Percent Hispanic}_i) + \beta_{10} (\text{Percent Asian Amerian}_i) + \epsilon_i
\end{multline*}

This model assumes that the impact of having a manager on turnout is the same across all elections. On the aggregate level this assumption clearly holds, because on average city managers typically have similar responsibilities and powers so should be correlated with the same change in turnout.

## Matching

City managers are not randomly assigned to municipalities. Instead each municipality chooses whether or not to adopt the manager form of government. This introduces confounding into the data, as the decision to have a manager is contingent on many factors such as population size, administrative needs, resident priorities, etc., which are also affecting turnout at the same time. Although OLS regressions accounts for this and adjusts coefficients accordingly, this adjustment can make it hard to use them as causal estimate and interpret properly. For example, if population size is highly correlated with having a manager, such that cities with and without managers have completely different populations as well, then the coefficient will give an estimated association of the manager on turnout holding population constant. However, if this comparison does not exist within data, the regression's estimate of this association will be biased.

To address this, I implemented a matching procedure through the `MatchIt` package. [@ho_matchit_2024]. This procedure involves reducing the dataset down to pairs of elections, for which the only difference between them is whether they have a city manager or not. Therefore, now the comparison of all else equal, does exist, creating the opportunity to have a less biased representation of the impact city managers have on turnout. Although perfect matching is impossible, the procedure is still able to remove samples that may be biasing the analysis of the regression. The regression model will then be more able to capture a more realistic representation of how city managers effect turnout, holding other factors constant. This procedure, however, does not allow for causal estimates, because the covariates I use are not determined pre-treatment. The demographic, institutional, and electoral competitive factors, have dramatically shifted since the initial switch to a city manager systems, and likely because of the manager. For example, if managers are correlated with better governance, and better outcomes for residents, than people will move to move to those cities, and elections might become less competitive. Therefore, they are still confounding the relationship because the data on this time is not present or available. Despite this, the procedure is still worthwhile for its ability to prune the dataset and produce a more reliable regression [@ho_matching_2007, see figure 1] with less bias coming from the measured covariates I have. All the additional covariates included in my linear model were used to match observations. Its also important to note that this procedure does not reduce bias across all the regression coefficients, but only for the coefficient for city managers.

The `MatchIt` procedure has two key parameters for tuning which are the matching algorithm and method for determining the distance between two observations. Determining the distance can be down via a propensity score or distance metric like euclidean distance, while the matching algorithm determines how observations are paired based these scores [see @greifer_matching_2024 for a full list of methods]. The authors [@greifer_matchit_2024] recommend balance metrics and cutoffs which determine the performance of each method combination. The 2 metrics used for this analysis are the standardized mean difference and the variance ratio between each group in the matched sample. The cutoff for for the standardized mean difference is recommended to be within -0.1 and 0.1 ($|{\overline{X_{iC}} – \overline{X_{iT}}}| < 0.1$), for each covariate, and the variance ratios between 0.5 and 2 ($0.5 \le \frac{Var(X_{iC})}{Var(X_{iT})} \le 2$). When these metrics are inside these cutoff ranges, it indicates that the distributions of each covariate is relatively for cities with managers and those without. If these conditions cannot be met, it means the matching is unsuccessful and likely not useful. After testing every combination of methods, the best results were produced by the optimal matching algorithm and logistic regression for propensity score calculations. This combination produced 10 balanced means out of 12 variables and 6 balanced variances out of 6 continuous variables.


This optimal matching method seeks to minimize the absolute sum of differences between propensity scores of each pair, and therefore tests every possible combination of pairs until this function is minimized. Then the matches for the minimized difference are returned as the selected dataset. A mathematical representation of this process is below:

\begin{multline*}
\text{argmin} \sum\limits_{i=1}^{\min(C,T)} \left| P_{Ci} - P_{Ti} \right|\\
\end{multline*}

The logistic regression is used to estimate the probability of each observation having a city manager based on the key covariates which is used as the propensity score. The functional form for the model:

\begin{multline*}
P(\text{Has a Manager}_i = 1 \mid X_{1i}, \dots, X_{10i}) = \\
\frac{1}{1 + e^{-( \beta_0 + \beta_1 X_{1i} + \cdots + \beta_{10} X_{10i})}}\\
\text{where:} \\
X_1 = \text{Concurrency, } X_2 = \text{Mayoral Government, } X_3 = \text{Contested Office, } \\
X_4 = \text{Incumbent Present, }  X_5 = \text{Margin of Victory, }  X_6 = \text{Percent Women, } \\
X_7 = \text{Percent Black, }  X_8 = \text{Percent Hispanic, }  X_9 = \text{Percent Asian, } X_{10} = \text{Population (ln)}
\end{multline*}

@tbl-bal and @fig-love showcase the success of this matching procedure with the matched sample scoring better on these key metrics than the unadjusted sample. @tbl-bal reports standardized mean difference and variance ratios both before and after matching, while also showcasing which covariates are balanced based on the criteria above. @fig-love shows the same visually using blue dots for the matched sample and red dots for the unmatched. Dots within the dashed lines are considered balanced. Overall based on the figure and metrics, the procedure was very successful and balancing the covariates in the sample. However unbalanced variables such as population size, and the percentage of Black residents are still present and therefore the confounders remaining among the observed data. The precise distributions before and after matching for each variable are in the appendix at @fig-bal. These figures were produced by `cobalt` [@greifer_cobalt_2024], an accompanying package to `MatchIt` used for assessing balance.

```{r}
#| output: 'asis'
#| label: tbl-bal
#| tbl-cap: "Balance Table Results for Matching Procedure"
final_balances <- bal.tab(optimal_glm, stats = c("mean.diffs", "variance.ratios"), 
                         binary = "std", thresholds = c(m=0.1, v=2), un = TRUE)
balance_table <- as_tibble(final_balances[[1]]) %>% 
  mutate(Covariates = c("Propensity Score", "Not Concurrent", 
                        "Concurrent With Midterm","Concurrent with President", 
                        "Has a Mayor",
                        "Mayoral Election", "Population (ln)",
                        "Incumbent Prescence", "Margin of Victory", "Women (%)",
                        "Black (%)", "Hispanic (%)", "Asian (%)")) %>%
  select(Covariates, Type, Diff.Un, V.Ratio.Un, Diff.Adj,
       M.Threshold, V.Ratio.Adj, V.Threshold) %>% mutate(
         Diff.Un = round(Diff.Un, 4),
         V.Ratio.Un = round(V.Ratio.Un, 4),
         Diff.Adj = round(Diff.Adj, 4),
         V.Ratio.Adj = round(V.Ratio.Adj, 4))
stargazer(balance_table, summary = FALSE, header = FALSE,
          digits = 4, column.sep.width = "0pt", single.row = TRUE,
          rownames = FALSE, font.size = "footnotesize", float = FALSE)
```

```{r}
#| fig.cap: "Love Plot for Matched Covariates"
#| fig.width: 11
#| fig.height: 15
#| label: fig-love
bal_names <- list("Propensity Score", "Not Concurrent", "Concurrent With Midterm",
                        "Concurrent with President", "Has a Mayor",
                        "Mayoral Election", "Population (ln)",
                        "Incumbent Prescence", "Margin of Victory", "Women (%)",
                        "Black (%)", "Hispanic (%)", "Asian (%)")
names(bal_names) <- var.names(final_balances)

love <- love.plot(optimal_glm, stats = c("mean.diffs", "variance.ratios"), 
          binary = "std", thresholds = c(m=0.1, v=2), un = TRUE,
          var.names = bal_names, title = NULL)
```
\newpage
This new matched sample has `r nrow(matched_data)` elections, with 97 pairs of elections. Summary Statistics for the matched sample are above in @tbl-summ, but some major differences are now that all cities have mayors (as opposed to `r round(mean(final_contests$has_mayor),1)*100`%), `r round(mean(matched_data$mayor_elec),2)*100`% of elections were mayoral (up from `r round(mean(final_contests$mayor_elec), 2)*100`%), only `r round(mean(matched_data$concurrent_pres),2)*100 + round(mean(matched_data$concurrent_mid),2)*100`% of elections are concurrent with presidential, midterm, or gubernatorial elections (down from `r round(mean(final_contests$concurrent_pres),2)*100 + round(mean(final_contests$concurrent_mid),2)*100`%), and average turnout has dropped from `r round(mean(final_contests$turnout),0)`% to `r round(mean(matched_data$turnout),0)`%, while standard deviation as decreased from `r round(sd(final_contests$turnout),0)`% to `r round(sd(matched_data$turnout),0)`%. While these variable differences are not that important substantively, its important to recognize that the shifting of these summary statistics is a direct result of the matching procedure working properly to remove pair together observations of similar values. Its clear that these differences are resulting from the wide ranging values in the original sample which made the two groups incomparable.

```{r}
#| output: 'asis'
#| tbl-cap: "Summary Statistics of Selected Covariates in Matched Sample"
#| label: tbl-summ
matched_data %>% select(turnout, council.manager, incumbent,margin, 
                          percent_women, percent_black , percent_hispanic , 
                          percent_asian_american, log.pop, concurrent_pres,
                          concurrent_mid,has_mayor,mayor_elec) %>%
  rename("Turnout(%)" = turnout,
         "Has a Manager" = council.manager,
         "Has a Mayor" = has_mayor, 
         "Mayoral Election" = mayor_elec, 
         Incumbency = incumbent,
         "Population(ln)" = log.pop,
         "Margin of Victory(%)" = margin,
         "Women(%)" = percent_women,
         "Black(%)" = percent_black,
         "Asian(%)" = percent_asian_american,
         "Hispanic(%)" = percent_hispanic,
         "Concurrent Presidential" = concurrent_pres,
         "Concurrent Midterm" =  concurrent_mid) %>%
  as.data.frame() %>%
  stargazer( summary = TRUE, header = FALSE, digits = 2,
          single.row = TRUE, column.sep.width = "0pt", font.size = "footnotesize",
          float = FALSE, omit.summary.stat = "N")
```

I conducted my analysis using R version 4.41 [@rcoreteam2024], while relying on several `tidyverse` packages [@wickham_welcome_2019], the RStudio IDE [@positteam2024] , `MatchIt` for matching [@ho_matchit_2024], `Cobalt` for assessing balance [@greifer_cobalt_2024], `Stargazer` for all the tables [@hlavac_stargazer_2022], and `gridExtra` for arranging plots [@auguie_gridextra_2017].

# Results

## Regression Analysis

```{r}

lm_ref <- lm(turnout ~ council.manager, final_contests)
lm_full <- lm(turnout ~ council.manager + concurrent + has_mayor + office +
                incumbent + margin + percent_women + percent_black + log.pop + 
                percent_hispanic + percent_asian_american, data = final_contests)

lm_ref_matched <- lm(turnout ~ council.manager, data = matched_data, weights = weights)
lm_full_matched <- lm(turnout ~ council.manager + concurrent+ office + 
                        has_mayor + incumbent + margin + log.pop + 
                        percent_women + percent_black + percent_hispanic +
                percent_asian_american, data = matched_data, weights = weights)


```

```{r}
models <- list(lm_ref,lm_full, lm_ref_matched, lm_full_matched)
p.values <- vector()

for (i in seq_along(models)) {
  tidied <- tidy(models[[i]])  %>% filter(term == "council.manager") %>% select(p.value)
  p.values[i] <- as.vector(tidied)
}
adjusted <- p.adjust(p.values, method = "BH")
```

@tbl-regression below, reports the results of the regression models with confidence intervals for the coefficients in parentheses, and $p$-values less than 0.05, are noted with asterisks. It is important to note that all the cities with managers have mayors, while all the cities without managers do not have mayors, meaning the covariate no longer influences the model. Additionally, the model’s residuals feature heteroskedasticity (see @fig-resid; the variance of the residuals increases with the fitted values.), violating one of the key assumptions of the linear model, making the estimates of standard errors and p-values flawed. This issue is less concerning given I’ve failed to reject my most important hypothesis. For all the hypothesis testing in this section $\alpha = 0.05$. The p-values for Manager coefficients have been adjusted using the “BH” method, to account for the higher type I error rate associated with multiple tests. This adjustment was made to balance the risk of increased type I errors and the power of the tests, although only 4 tests were conducted. The adjusted p-values do not change any of the rejection decisions.

```{r}
#| output: 'asis'
#| label: tbl-regression
#| tbl-cap: "Regression Model Results: With 95% Confidence Intervals"

stargazer(lm_ref, lm_full, lm_ref_matched, lm_full_matched,
          type= "latex", header = FALSE, ci = TRUE,
          ci.level = 0.95,
          df = TRUE, font.size = "footnotesize", 
          model.names = TRUE,
          column.sep.width = "0pt", digits = 2,
          star.cutoffs = c(0.05, NA, NA), notes = " *p<0.05", 
          notes.append = FALSE,
          covariate.labels = c("Manager", "Midterm Concurrency", "President Concurrency",
          "Has a Mayor", "Mayoral Election", "Incumbent Presence", "Margin of Victory",
          "Percent Women", "Percent Black", "Population(ln)", "Percent Hispanic", "Percent Asian"),
          dep.var.labels = "Voter Turnout",
          no.space = TRUE,
          single.row = TRUE,
          column.labels = c("Unmatched", "Matched"),
          column.separate = c(2,2), float = FALSE)


```

In the first model, City Managers are associated with a 9.48% increase in turnout compared to cities without managers, with a 95% confidence interval from 4.47% to 14.50%. However, accounting for other covariates in the second model reverses this association to a 3.59% decrease in turnout, with a 95% confidence interval from -6.57% to -0.60%. The changes in the coefficient indicate confounding in the data, as described earlier, which the model attempts to account for. In the models using the matched data, initially, city managers (without covariates) are associated with a 2.72% decrease in turnout, with a 95% confidence interval between -7.65% and 2.21%, while after accounting for covariates city managers are associated with a 2.54% decrease in turnout, with a 95% confidence interval between -5.23% and 0.16%. The much smaller change between the third and fourth models is another indicator that the matching procedure was successful. If the covariates were still drastically confounding the sample, then it would have changed the coefficient for managers as occurred between models one and two.

The fourth model presents my final estimate of the relationship between city managers, which is -2.54%, meaning that when all other factors are held constant, I would expect turnout to be 2.54% lower in a city with a manager. However, as mentioned before the procedure does not ensure a causal interpretation because many unmeasured confounders are still impacting the results. For example, I lack data on institutional characteristics related to the intricacies of power such as term limits and length for officials, economic status, and electoral system each city uses. These factors were used in previous research[@hajnal_municipal_2003;@caren_big_2007] and thus excluding them willingly even with the lack of data introduces bias. Additionally, an important unmeasured factor is the state legislative environment which is completely variable. Each state has its own rules regarding voter registration, city charters (the documents that define a municipality's scope and authority), along with the types of governments municipalities are allowed to have. All of these factors have the potential be correlated with turnout rate, and whether or not a city has a manager, so their lack of inclusion biases the results even with the matching. 

The matching has only clarified the relationship to the regression model, given what covariates I include from the data. Nevertheless the level of bias for city manager coefficient has still been reduced, providing a more accurate estimate for the true relationship. The 95% confidence interval (-5.23%, 0.16%), around this coefficient indicates the general degree of uncertainty. If I were to collect 100 other samples of elections and create similar 95% confidence intervals, I would expect 95% of these intervals to contain the true underlying value which is unknown. The actual width of the interval reflects the variance of the coefficient, as when the variance is lower, or the sample size is larger we can be similarly confident with a smaller interval. Thus the constricting of the interval’s size (from around `r round(14.5 - 4.47, 1)`% to `r round(5.23 + 0.16, 1)`%) through the models, shows that the coefficient’s variance has drastically reduced even as there are `r nrow(final_contests) - nrow(matched_data)` fewer elections in the sample after matching.

## Hypothesis Testing

Testing these coefficients for significance, in all cases, my null hypothesis is that there is no relationship between city managers and turnout ($H_0: \beta_1 = 0$), while the alternate hypothesis, is that the relationship between city managers and turnout is not 0 ($H_A: \beta_1 \neq 0$). Setting $\alpha = 0.05$:

<div>

1.  In model 1, $p =$ `r options(scipen = 999); round(adjusted[1], 4)` and since $p < 0.05$, I reject the null hypothesis.
2.  In model 2, $p =$ `r round(adjusted[2], 4)` and since $p < 0.05$, I reject the null hypothesis.
3.  In model 3, $p =$ `r round(adjusted[3], 4)` and since $p > 0.05$, I fail to reject the null hypothesis.
4.  In model 4, $p =$ `r round(adjusted[4], 4)` and since $p > 0.05$, I fail to reject the null hypothesis.

</div>

These p-values represent the probability of seeing the coefficients of that magnitude given the $H_0$ that the true $\beta_1 =  0$. It is a measure of how extreme the result is compared to what I assume from the $H_0$. When these probabilities are lower than my set $\alpha$, I have enough evidence to reject $H_0$'s assumption, but when they are higher than $\alpha$ I do not, and therefore fail to reject $H_0$. Failing to reject, suggests that the obtained value is consistent with a hypothetical sampling distribution of $\beta_1$'s based on the null hypothesis' assumption.

Using the fourth model’s coefficient as my final estimate of the relationship between city managers and voter turnout in municipal elections, my failure to reject the null hypothesis indicates my results are not statistically significant. Therefore, I cannot conclusively say that city managers have any relationship without a turnout rate, and these observed results could be due to random noise. This directly contradicts previous research that has documented the relationship as significant [@hajnal_municipal_2003; @caren_big_2007; @wood_voter_2002], and their substantive interpretations of their results. Based on my sample, I cannot predict with certainty that removing a city manager would result in any change in turnout.

# Discussion

The goal of this study was to isolate the impact of city managers on turnout, in hopes of identifying a simple, effective solution for cities to implement. The small magnitude of the coefficient (only around 2.5%) suggests that even if this could be estimated as a causal effect, the real-world impact of switching forms of government would be marginal. Furthermore, the lack of statistical significance indicates cities struggling to increase turnout in their local elections should consider other policy options instead.

This information is still useful: professional city managers can potentially run cities more efficiently, improving access to many of the essential services voters would care about. If this is true, then it’s clear why the relationship is relatively benign compared to other factors in the model such as concurrency with midterm and presidential elections. The perceived lack of power that may arise could be balanced by the potential for a municipal government that is more responsive to the residents needs and desires. If raising turnout would come at the expense of any benefits provided by a manager, it would be hard to justify the change.

Although its certainly possible to imagine an individual voting because their perceived lack of influence this study has shown at the aggregate level the relationship city managers and turnout is not statistically significant. It is certainly worth considering that this perception of city managers may no longer hold for contemporary voters who are so far removed from the initial implementation of these reforms they are unaware of their city manager.
Its clear from the regression that the most influential variable is concurrency with the presidential election and midterms (associated with 30% and 24% respective increases in turnout in model 4), indicating that the importance of the federal election is what draws voters to the polls. Therefore the people who turnout out in non federal election years are uniquely committed to participating in local affairs and will likely participate no matter the governance structure. Conversely, the people who only vote in federal election year, are clearly not committed to local politics, and whether a city manager can change that is unknown but unlikely given the results of the study. Future studies into local politics should investigate this relationship, and perhaps focus more on what causes an individual to turnout instead of what is associated with the aggregate turnout rate.

This study has the potential to raise important concerns within the field by contradicting previous research. However, it is not without limitations. Most importantly, the use of matching can never replicate randomized experiments. The distributions of each variable are still not perfectly aligned after matching (see @fig-bal). This means all even the measured covariates still potential confounders, with the unbalanced covariates being the most salient (population size, and percentage of black residents). Additionally, the matching procedure can only work on measured characteristics so unmeasured confounders or those not adequately measured in the data such as the competitiveness of an election, the power of an office, or the overall political culture of a region are still impacting the results. Additionally, the reduction in sample size also poses a problem, as it becomes more difficult to generalize the results. Verifying the results of this paper involves leveraging the rest of this electoral database [@benedictis-kessner_american_2023] with similar matching methods to see if these findings hold. Sampling from the dataset also had drawbacks. Random sampling ensures a representative sample from the population, but if that population is not representative than its also flawed. The cities that ended in the sample do not appear to be representative of the nation at large. (see @fig-map). Many cities are concentrated in California, and Connecticut, with hardly any in the Northwest and Southeast. This is far more diverse than what other recent researchers have accomplished but still falls short.

By no means do these limitations render this study irrelevant but the caveats with methodology and interpretation are paramount to consider when using results to influence policy outcomes. These results are most readily applicable to the cities contained within the sample, which the results directly reflect. However they can also be generalized (carefully) to any municipality that broadly has similar characteristics and is located in a similar region to a sampled city. More importantly, is that this work contributes to an area of research that is sparse and underdeveloped, while also applying new methods to provide a unique take on the issue. Additionally, by diverging from other researchers in the field [@wood_voter_2002; @hajnal_municipal_2003; @caren_big_2007] this work highlights an  unclear consensus in the field, showing that more discussion is needed. The findings presented by this paper serve as a stepping stone to a better understanding of municipal governance and local political participation in the United States.

Ideally, the next steps for future research would be to replicate the analyses conducted by Caren [@caren_big_2007] and Hajnal and Lewis [@hajnal_municipal_2003], on all the cities in this electoral database, using the same detailed information, providing more evidence on all the correlates of turnout. However, for a robust study of city managers, the next steps would require gathering important pre-manager municipal data so that more robust causal estimates could be made from regressions or difference-in-difference design.
\newpage

# Appendix {.appendix}

```{r}
#| output: 'asis'
#| tbl-cap: "List of Sampled Cities"
#| label: tbl-city

city <- final_contests %>% group_by(city) %>% reframe(city = city, state = state) %>%
  ungroup() %>% filter(!duplicated(city))

stargazer(city, summary = FALSE, header = FALSE, type = "latex", 
          font.size = "tiny", column.sep.width = "0pt",
          float = FALSE)
```

```{r}
#| fig.cap: "Map of Sampled Cities"
#| fig.width: 7
#| label: fig-map
places <- read_csv("../data/places_geocoded.csv")
places$fips <- as.numeric(places$fips)
sample$geo_name <- tolower(sample$geo_name)
sampled_places <- final_contests %>%
  group_by(city) %>% reframe(geo_name = city,
                             state = state) %>% 
  ungroup() %>%
  filter(!duplicated(geo_name)) %>%
  left_join(sample, by = c("geo_name")) %>%
  left_join(places, by = "fips")
usa <- map_data("usa")
ggplot()+
  geom_map(map = usa)+
  borders(database = "state")+
  geom_point(aes(lng, lat),data = sampled_places, shape = 2)+
  theme_minimal()+
  labs(x = NULL, y = NULL)+
  theme(legend.position = "none")+
  scale_x_continuous(breaks = NULL)+
  scale_y_continuous(breaks = NULL)

```

```{r}

all_covs <- final_contests[, c("council.manager","turnout", "has_mayor", "mayor_elec", "incumbent", "log.pop", 
                       "margin", "percent_men", "percent_women", "percent_white",
                       "percent_black", "percent_hispanic", "percent_asian_american",
                       "concurrent_pres", "concurrent_mid")]
names(all_covs) <-c("Has a Manager", "Turnout", "Has a Mayor", "Mayor Election", 
                 "Incumbency", "Population(ln)", "Margin", "Men(%)", "Women(%)",
                 "White(%)", "Black(%)", "Hispanic(%)", "Asian(%)", "Concurrent Pres.",
                 "Concurrent Mid.")
inst_cov <- all_covs[, c("Has a Manager", "Turnout", "Has a Mayor", "Mayor Election",
                         "Incumbency", "Margin")]
dem_covs <- all_covs[, c("Has a Manager", "Turnout", "Population(ln)", "Men(%)", "Women(%)",
                 "White(%)", "Black(%)", "Hispanic(%)", "Asian(%)")]
tim_covs <- all_covs[,c("Has a Manager", "Turnout", "Concurrent Pres.","Concurrent Mid.")]

inst_cors<- as.data.frame((cor(inst_cov)))
dem_cors <- as.data.frame((cor(dem_covs)))
tim_cors <- as.data.frame((cor(tim_covs)))
```

```{r}
#| output: 'asis'
#| label: tbl-icor
#| tbl-cap: "Correlation Matrix of Institutional Covariates"
stargazer(inst_cors, summary = FALSE,  
          header = FALSE,
          digits = 2, single.row = TRUE,
          column.sep.width = "0pt", font.size = "footnotesize",
          float = FALSE)
```

```{r}
#| output: 'asis'
#| label: tbl-dcor
#| tbl-cap: "Correlation Matrix of Demographic Covariates"
stargazer(dem_cors, summary = FALSE, 
          header = FALSE,
          digits = 2, single.row = TRUE,
          column.sep.width = "0pt", font.size = "scriptsize",
          float = FALSE)
```

```{r}
#| output: 'asis'
#| tbl-cap: "Correlation Matrix of Timing Covariates"
#| label: tbl-tcor
stargazer(tim_cors, summary = FALSE, 
          header = FALSE,
          digits = 2, single.row = TRUE,
          column.sep.width = "0pt", font.size = "footnotesize",
          label = "tbl-tcor", float = FALSE)
```

```{r}
#| fig.cap: "Individual Distributional Balance for Matched Covariates"
#| fig.width: 10
#| fig.height: 7
#| label: fig-bal
bals <- c("incumbent", "percent_women", "percent_black", "percent_hispanic",
             "percent_asian_american", "log.pop", "concurrent", 
          "has_mayor", "distance")
bal_labs <- c("Incumbent", "Women (%)", "Black (%)", "Hispanic (%)", "Asian (%)", 
              "Population (ln)", "Election Concurrency", "Has a Mayor",
              "Propensity Score")

bal_plots <- list()


for(i in seq_along(bals)) {
  bal_plots[[i]] <- bal.plot(optimal_glm, var.name = bals[i], which = "both",
                             sample.names = c("Unmatched", "Matched"))+
    ggtitle(NULL) +
    labs(
      x = paste(bal_labs[i]), 
      y = ifelse(
        bal_labs[i] %in% c("Incumbent", "Election Concurrency", "Has a Mayor"), 
        "Proportion", 
        "Density")
    ) +
    theme(
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks = element_blank()
    )
}
grid.arrange(grobs = bal_plots, nrow = 3)
```

```{r}
#| fig.cap: "Residual Plots for the Multiple Regression Models"
#| label: fig-resid
#| fig.height: 3
#| fig.width: 5
resid_plot_un <- ggplot(augment(lm_full), aes(.fitted, .resid))+
  geom_point() + 
  geom_hline(yintercept = 0, linetype = 2, color = "blue")+
  labs(title = "Unmatched", x = "Fitted Values", y = "Residuals")+
  theme_minimal()+
  ylim(c(-50,50))
resid_plot_ma <-ggplot(augment(lm_full_matched),
                                aes(.fitted, .resid))+
  geom_point() + 
  geom_hline(yintercept = 0, linetype = 2, color = "blue")+
  labs(title = "Matched", x = "Fitted Values", y = "Residuals")+
  theme_minimal()+
  ylim(c(-50,50))
grid.arrange(resid_plot_un, resid_plot_ma, ncol = 2)
```

# References {.appendix}
